\section{Conclusion}

ALNS consistently outperforms Tabu Search. As both algorithms reaches a point where it becomes hard to find a much better solution, this must be because ALNS covers a much broader solution space.

Diversification in Tabu Search is meant to move the search to another part of the solution space. But clearly this is not accomplished sufficiently. This is likely because Tabu Search spends a lot of time just in validating solutions, this is particularly the case with the swap operation. ALNS don't do this, as it just looks at the best placement for missing courses. Improving the performance of the move and swap operations could make a huge difference for Tabu Search. The fact that the dynamic neighborhood consistently outperforms any other neighborhood credits this hypothesis too.

Even if the move and swap operations where made faster, it is possible that the complexity of the problem makes it impossible for those operations to become sufficiently fast. Using a dynamic neighborhood thus still makes sense. In particularly this is something that could be applied to many other problems. It has definitely been shown to make a big improvement.

In parameter tuning for ALNS it was found that using the same $w_{current}$ and $w_{global}$ gave the best results. This is something that is not intuitive at first, but as discussed indicates that a globally better solution is found by a long line of good repair and destroy choices. If one did not have the computational resources to optimize these parameters, setting them to the same value may be a good choice.

To improve the ALNS method further, one could analyze the repair and destroy method selection probabilities to learn about what works and more importantly doesn't work for some datasets. Using this knowledge more fine tuned algorithms could be constructed. However one also have to be careful not to do such must fine tuning that it becomes overfitting.

At last one could explore combining the methods. A simple way of doing this, could be to initialize the Tabu Search using the ALNS solution. This would ensure that the ALNS solution is in a local minima and there isn't some nearby better solution, that can be found by simple diversification.
